# Train a Random Forest {#sec-modelfit}

In this chapter, we are going to use the data prepared in @sec-dataprep, train a Random Forest model, and evaluate the variable importance of the different predictors used in the model. To recap the basics of the Random Forest algorithm and its implementation and use in R, head over to [Chapter 11 of AGDS book](https://geco-bern.github.io/agds/randomforest.html).

## Load data

In the previous Chapter, we created a dataframe that holds information on the soil sampling locations and the covariates that we extracted for these positions. Let's load this datafarme into our environment

```{r}
data_clean <- readRDS(here::here("data/bern_sampling_locations_with_covariates.rds"))

head(data_clean) |> knitr::kable()
```

## Preparations

Before we can fit the model, we have to specify a few settings. First, we have to specify our response and predictor variables. Then, we have to split our dataset into a calibration and a validation set. Random Forest models cannot deal with `NA` values, so we have to remove these from our calibration set. 

In the dataset we work with here, the data splitting into training and testing sets is defined by the column `dataset`. Usually, this is not the case and we split the data ourselves. Have a look at the [section in AGDS Book on data splitting](https://geco-bern.github.io/agds/supervisedmli.html#data-splitting).

As predictors we use all the variables for which we have data with spatial coverage - the basis for spatial upscaling. We extracted these data in @sec-dataprep from geospatial files and the data frame was constructed by `cbind()`, where columns number 14-28 contain the covariates data, extracted from the geospatial files. We use these here as predictors for the model.

```{r}
# Specify response: The pH in the top 10cm
response <- "ph.0.10"

# Specify predictors: Remove soil sampling information
predictors <- names(data_clean)[14:28]

cat("The response is:", response,
    "\nThe predictors are:", paste0(predictors[1:8], sep = ", "), "...")
```

```{r}
# Split dataset into calibration and validation
data_cal <- data_clean |> dplyr::filter(dataset == "calibration")
data_val <- data_clean |> dplyr::filter(dataset == "validation")

# Filter out any NA to avoid error when running a Random Forest
data_cal <- data_cal |> tidyr::drop_na()
data_val <- data_val |> tidyr::drop_na()

# A little bit of verbose output:
n_tot <- nrow(data_cal) + nrow(data_val)

perc_cal <- (nrow(data_cal) / n_tot) |> round(2) * 100
perc_val <- (nrow(data_val) / n_tot) |> round(2) * 100

cat("For model training, we have a calibration / validation split of: ",
    perc_cal, "/", perc_val, "%")
```

Alright, this looks all good. We have our response and predictor variables saved for easy access later on and the 75/25% split of calibration and validation data looks good, too. We can now move on to model fitting.

## Model training

The modelling task is to predict the soil pH in the top 10 cm. Let's start using the default hyperparameters used by `ranger::ranger()`.

::: callout-tip
Have a look at the values of the defaults by entering. `?ranger::ranger` in your console and study the function documentation. 
:::

```{r}
# ranger() crashes when using tibbles, so we are using the
# base R notation to enter the data

rf_basic <- ranger::ranger( 
  y = data_cal[, response],   # Response variable
  x = data_cal[, predictors], # Predictor variables
  seed = 42,                  # Specify the seed for randomization to reproduce the same model again
  num.threads = parallel::detectCores() - 1) # Use all but one CPU core for quick model training

# Print a summary of fitted model
rf_basic |> print()
```

::: callout-tip
## Predicting categories with Random Forests

If our response variable was a categorical and not a continuous variable, we would have to set the argument `probability = TRUE`. The output would then be a probability map from 0-100%.
:::

Although we only used the pre-defined parameters, we already get a fairly good out-of-bag (OOB) $R^2$ of 0.41 and a MSE of 0.35 pH units. 

<!-- ### Model with weights -->

<!-- Sometimes we know that a subset of our dataset is more trustworthy than the rest. For example, when you are using a gap-filling technique to interpolate data, that gap-filled data is less trustworthy than the actually measured data. Informing the model algorithm that it should weigh certain data entries more than others can affect the final model performance. -->

<!-- In our dataset, we have information on whether the pH values were measured in the field - which is less precise - or in the lab. Also, two different lab methods are indicated in the data. All of this information is held in the suffix of the `timeset` variable, so let's assign weights according on the quality of the pH data as follows: `1` for CaCl$_2$ lab measurement (no suffix), `0.9` for pedotransfer from another lab method (suffix `_ptf`), and `0.7` for field data (suffix `_field`). For this, we create a weight-matching vector: -->

<!-- ```{r} -->
<!-- weights <-  -->
<!--   data_cal |> -->
<!--   dplyr::mutate( -->
<!--     # Create a new variable 'weight' which holds only 1's -->
<!--     w = 1, -->
<!--     # Check the suffix in each row and if true, give a new weight. If false, keep the old weight. -->
<!--     w = ifelse(stringr::str_detect(timeset, "_field"), 0.9, w), -->
<!--     w = ifelse(stringr::str_detect(timeset, "_ptf"), 0.7, w) -->
<!--   ) -->

<!-- # Quality check if everything worked: -->
<!-- set.seed(42) -->

<!-- weights |>  -->
<!--   dplyr::select(timeset, w) |>  -->
<!--   dplyr::slice_sample(n = 8) |>   # Pick 8 random rows -->
<!--   knitr::kable() -->
<!-- ``` -->

<!-- It is always a good idea to do quality checks when wrangling! Here we see that our weight attribution code worked as expected, so we can move on to model fitting. -->

<!-- ```{r} -->
<!-- rf_weighted <- ranger::ranger(  -->
<!--   y = data_cal[, response],      # Response variable -->
<!--   x = data_cal[, predictors],    # Predictor variables -->
<!--   case.weights = weights[, "w"], # Add weights to input -->
<!--   seed = 42,                     # Specify seed for randomization to reproduce the same model again -->
<!--   num.threads = parallel::detectCores() - 1) # Use all but one CPU core for quick model training -->

<!-- # Print a summary of fitted model -->
<!-- rf_weighted |> print() -->
<!-- ``` -->

<!-- Not much has changed compared to our previous model. We see that the $R^2$ and MSE got negligibly worse but as a trade-off we gained more trust in our model. -->

This is the step at which you may want to optimize the hyperparameters for improving the model performance and generalisability. You may want to compare  different hyperparameter specifications of the Random Forest model that control the model complexity. A simple way to do that is to use the {caret} R package which provides machine learning wrapper functions for hyperparameter tuning (among many more functionalities). Its use in combination with Random Forest is demonstrated in [Chapter 11 of AGDS book](https://geco-bern.github.io/agds/randomforest.html).

## Model interpretation

### Variable importance based on OOB-Error

Our model has 15 variables, but we have no idea if each of them should really be in the model and if we are not just fitting noise in the dataset. To investigate this issue, the `ranger()` function takes an argument called `importance`. We can set this argument either to follow the `permutation` method, whereby the algorithm randomly permutes values of each variable and measures the resulting decrease in model accuracy. A larger decrease indicates a more important variable. If the code runs slow, you can also use the faster `impurity` method (see more information [here](https://bmcbioinformatics.biomedcentral.com/counter/pdf/10.1186/s12859-016-0995-8.pdf)).

Assessing the variable importance gives us a feeling for what variables we should keep or drop from the dataset. The ranger-model stores this information if we enter a `importance` method. The code below accesses the model's variable importance and sorts the variables with decreasing importance.

```{r fig.height=6, fig.width=12}
# Let's run the basic model again but with recording the variable importance
rf_basic <- ranger::ranger( 
  y = data_cal[, response],     # Response variable
  x = data_cal[, predictors],   # Predictor variables
  importance   = "permutation", # Pick permutation to calculate variable importance
  seed = 42,                    # Specify seed for randomization to reproduce the same model again
  num.threads = parallel::detectCores() - 1) # Use all but one CPU core for quick model training

# Extract the variable importance and create a long tibble
vi_rf_basic <- 
  rf_basic$variable.importance |>
  dplyr::bind_rows() |> 
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "variable")

# Plot variable importance, ordered by decreasing value
p <- 
  vi_rf_basic |> 
  ggplot2::ggplot(ggplot2::aes(x = reorder(variable, value), y = value)) +
  ggplot2::geom_bar(stat = "identity", fill = "grey50", width = 0.75) + 
  ggplot2::labs(
    y = "Change in OOB MSE after permutation", 
    x = "",
    title = "Change in OOB MSE after permutation") +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1))

# Display plot
p
```

What do we see here? The y-axis shows the decrease in model performance when the respective variable is randomly permuted and, therefore, denotes the importance of a variable. The higher the value, the stronger the effect of permutation on the model performance, the more important the variable. In our randomly picked subset of covariates, we see that the time of sampling, the location (x and y coordinates), and the mean annual precipitation are very strong predictors of soil pH. From a soil-forming perspective, this seems plausible. For example, soil weathering is strongly driven by water. So, the more water a location receives, the more weather a soil is, the more likely it has a low pH. To dig deeper into understanding how the model works, we could further investigate its partial dependence plos.

We can also see that a large part of our covariates have practically no power to predict the pH and can therefore be removed from the model. But how do we determine what variable to pick for our final model? Do we want a maximum number of predictors? Do we want a set of the top n predictors that make up a certain percentage of the total variable importance?

### Boruta Algorithm

One common option to generalize variable selection is the "Boruta-Algorithm", which itself is based Random Forests. In essence, the algorithm creates "shadows" of your original data, where each of the predictor value is randomly permuted, which destroys the predictive power of the variable. Then, the algorithm iterates over these "shadows" and assess for each variable whether its permutation has a substantial effect on the model performance or not. E.g., if a model trained on a variable's shadow performs constantly worse than when trained on the original values, that variable is assessed as important. The algorithm categorizes all variables into "to reject", "to consider", and "to keep". Luckily, we do not have to write this algorithm ourselves but can use the {Boruta} package:

```{r fig.height=8, fig.width=12}
set.seed(42)
bor <- 
  Boruta::Boruta(
    y = data_cal[, response], 
    x = data_cal[, predictors],
    maxRuns = 50, # Number of iterations. Set to 30 or lower if it takes too long
    num.threads = parallel::detectCores()-1)

# Plot variable importance, the Boruta-output can be directly fed into base R plot()
par(oma = c(8,3,2,2)) # enlarge plot below for long variable labels
plot(bor, 
     xlab = "", 
     ylab = "Importance",
     las = 2,
     )
```

```{r}
# Check whether the most important variables from Boruta-Algorithm are similar as the
# important variables from the weighted Random Forest model
bor_top10 <- 
  Boruta::attStats(bor) |> 
  tibble::rownames_to_column() |> 
  dplyr::arrange(dplyr::desc(meanImp)) |> 
  dplyr::slice_head(n = 10) |> 
  dplyr::pull(rowname)

vi_top10 <- 
  vi_rf_basic |> 
  dplyr::slice_head(n = 10) |> 
  dplyr::pull(variable)

cbind(vi_top10, bor_top10) |> 
  knitr::kable(col.names = c("RF Top 10", "Boruta Top 10"))
```

We see that apart from the `timeset` variable, the variable importance calculated by the Boruta-Algorithm differs quite a bit, compared to the simple variable importance assessment built into the `ranger()` function. To move forward, we could keep only variables that were classified as "to keep" or "to consider" by the Boruta-Algorithm:

```{r}
predictors_bor <- 
  bor$finalDecision |> 
  as.data.frame(nm = "assessment") |> 
  tibble::rownames_to_column(var = "variable") |> 
  dplyr::filter(assessment != "Rejected") |> 
  dplyr::pull(variable)
```

For this tutorial, we are happy to continue with our most basic Random Forest and save all necessary data as shown below. However, as shown with the model using weights, with the variable selection of the Boruta method, and with applying hyper-parameter tuning and cross-validation for a more robust model, there are many options to create a better model.

```{r}
# Save relevant data for model testing in the next chapter.
# Pick rf to save:
rf <- rf_basic

# Extract predictors used in this rf
predictors <- rf$forest$independent.variable.names

saveRDS(rf_basic,                   
        here::here("data/rf_for_ph0-10.rds"))

saveRDS(data_cal[, c(response, predictors)],
        here::here("data/cal_for_ph0-10.rds"))

saveRDS(data_val[, c(response, predictors)],
        here::here("data/val_for_ph0-10.rds"))
```
